Diverse Genotype-by-Weather Interactions in Switchgrass

### 2024-05-22 manuscript/revision2/SI_Appendix/SI_Appendix.qmd
revision: small edits to Section S1 for clarity

Type or paste text here.

@@

\dataset{dataset_two.txt}{SNP-associated effects and standard errors for the start of vegetative growth jointly re-estimated in the Midwest genetic subpopulation.}

\dataset{dataset_three.txt}{SNP-associated effects and standard errors for the start of vegetative growth jointly re-estimated in both Midwest and Gulf genetic subpopulations.}

\dataset{dataset_four.txt}{SNP-associated effects and standard errors for the start of reproductive growth jointly re-estimated in the Gulf genetic subpopulation.}

\dataset{dataset_five.txt}{SNP-associated effects and standard errors for the start of reproductive growth jointly re-estimated in the Midwest genetic subpopulation.}

\dataset{dataset_six.txt}{SNP-associated effects and standard errors for the start of reproductive growth jointly re-estimated in both the Midwest and Gulf genetic subpopulations.}

>>>>

We specified three qualitative categories of covariance structures across gardens: "canonical" covariance, with simple patterns of effect size covariance introduced in the initial *mash* manuscript; "data-driven" covariances derived from common patterns of SNP effects observed in the data and introduced in the initial *mash* manuscript, and "GxWeather" covariance, estimated from the covariance of empirical weather patterns at each garden over specific time frames before the phenological event.

Canonical covariance matrices can be created by *mash* and fall into four groups: an identity matrix, an equal effects matrix, singleton matrices, and simple heterozygousity matrices. The identity matrix has values of zero for covariance between different gardens and correlations of one within gardens, and represents the situation in which the effects in different conditions are independent. The equal effects matrix has values of one for all matrix elements, and represents identical effects across all conditions. Each singleton matrix has a value of one for a single garden and zeros for all other elements, representing a garden-specific effect. Three simple heterozygousity matrices are generated, which have covariances of 0.25, 0.5, or 0.75 for all off-diagonal matrix elements, and correlations of one for diagonal entries, representing moderately positively correlated effects across all sites.

Data-driven covariance matrices can also be created by *mash* and fall into two groups: an overall pattern of covariance between all conditions, denoted as $tPCA$, and the first five principle components of a singular value decomposition of the $tPCA$, denoted as `PCA_1`$PCA_1$ through $PCA_5$[@Urbut2019].

To create the GxWeather covariance matrices we transformed each genotype's green-up and flowering date at each garden to create date-related values that were summary functions for a weather variable across a specified date range. For example, the cumulative sum of the rainfall that occurred seven days prior to the flowering date was used to create one possible GxWeather cue variable for flowering date. The code for these calculations can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe/R/Analysis_v1.2_mash_using_greedy_covar.qmd, starting at line 208.

We defined both phenological dates as functions of five weather variables and eight to ten time frames (Table 1): Cumulative growing degree days (GDD) variables for each time frame were calculated as $GDD = \sum^{DATE}_{DATE - i} max(T_{mean} - T_{base}, 0)$, where $T_{mean}$ is the daily average temperature, defined as $(T_{max} + T_{min}) / 2$, $T_{base}$ is the base temperature of 12°C for switchgrass, $T_{max}$ is the maximum daily temperature, $T_{min}$ is the minimum daily temperature, $DATE$ is the phenological date, and $i$ is the time frame, or the number of days prior to the phenological date that GDD is summed across [@kiniry2005; @behrman2013]. Second, cumulative rainfall values in each time frame were calculated as $C_{rain} = \sum^{DATE}_{DATE - i} PRCP$, where PRCP is the daily rainfall measured in millimeters, $DATE$ is the phenological date, and $i$ is the time frame, or the number of days prior to the phenological date that rainfall is summed across. Third, day length in hours was determined for each time frame as a specific single day prior to the phenological date, as indicated by the time frame $i$, e.g. $DATE - i$. Day length was calculated as a function of latitude and day of the year as in [@forsythe1995]. Fourth, day length change in seconds was determined for a specific single day prior to the phenological date as indicated by the time frame, as $DYLN_{change (s)} = (DYLN_{DATE - i} - DYLN_{DATE-i-1})*60*60$, where $DYLN$ in the day length in hours, $DATE$ is the phenological date, and $i$ is the number of days prior to the phenological date. Fifth, average temperature in each time frame was defined as $T_{ave} = \sum^{DATE}_{DATE - i} (T_{max} + T_{min}) / 2$, where $T_{max}$ is the maximum daily temperature, $T_{min}$ is the minimum daily temperature, $DATE$ is the phenological date, and $i$ is the time frame, or the number of days prior to the phenological date that average temperature is summed across.

We then generated GxWeather covariance matrices derived from correlations in these GxWeather cue variables for the three subpopulations. These covariance matrices represent the correlations between genotypes for these weather-related variables across our eight common gardens. To create these matrices, we determined the correlations between these weather-related cue variables for identical genotypes grown at different common gardens and used these correlations to fill the off-diagonals of a eight-by-eight common garden matrix. For the diagonal of these matrices, we used the narrow-sense heritability for each weather-related cue variable at each garden, calculated using the same methods as Section S3. Matrices need to be positive semi-definite to be used in *mash*, so when the resultant matrices were not positive semi-definite, the `Matrix::nearPD` function was used to compute the nearest positive definite matrix to the correlation matrix [@Matrix]. Finally, all covariance matrices are rescaled so that the maximum diagonal element of the rescaled matrix is 1, as in [@Urbut2019]. Code used to generate these matrices can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe.git/R/Analysis_v1.2_mash_using_greedy_covar.qmd, starting at line 536.

We defined distinct sets of weather-based covariance matrices for the two phenological dates and for the three genetic subgroups. We used the same five weather variables to transform both green-up and flowering date, but defined the weather-based covariance matrices relative to each of the two phenological dates. Weather-based covariance matrices were also defined separately for individual genotypes from the Gulf, Midwest, and Both subpopulations.

@@

\dataset{Dataset_2_Midwest_vegetative_growth_date.csv}{SNP-associated effects and standard errors for the start of vegetative growth jointly re-estimated in the Midwest genetic subpopulation.}

\dataset{Dataset_3_Both_subpopulations_vegetative_growth_date.csv}{SNP-associated effects and standard errors for the start of vegetative growth jointly re-estimated in both the Midwest and Gulf genetic subpopulations.}

\dataset{Dataset_4_Gulf_reproductive_growth_date.csv}{SNP-associated effects and standard errors for the start of reproductive growth jointly re-estimated in the Gulf genetic subpopulation.}

\dataset{Dataset_5_Midwest_reproductive_growth_date.csv}{SNP-associated effects and standard errors for the start of reproductive growth jointly re-estimated in the Midwest genetic subpopulation.}

\dataset{Dataset_6_Both_subpopulations_reproductive_growth_date.csv}{SNP-associated effects and standard errors for the start of reproductive growth jointly re-estimated in both the Midwest and Gulf genetic subpopulations.}
